---
title: "Canonical correspondence analysis for winter controls on phytoplankton distribution"
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    number_sections: yes
    pandoc_args: [
      "-V", "classoption=threecolumn"
      ]
    fig_caption: yes
    toc_float: yes
    theme: readable
---


<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}

</style>

<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
    word-break: break-word;
  }
  .col5 {
    columns: 5 80px;
    -webkit-columns: 5 80px;
    -moz-columns: 5 80px;
  }
</style>

<style>
div.tbcol {
  width: 500px;
}
</style>

# Introduction

This document presents a multivariate analysis of the relationships between trace metals and phytoplankton community composition. It provides a brief overview of the stastical methods and interpretation. This is followed by the code and outputs.

# Method overview

The factors influencing the phytoplankton community structure were investigated using a canonical correspondence analysis (CCA). The analysis was conducted in R using the `cca` function from the Vegan package. The biotic variables were represented by the CHEMTAX-derived taxonomical groups' biomass (µg/L of Chl a). The CCA was repeated for multiple sets of environmental parameters:

 * temperature, salinity, macronutrients, trace metals (dissolved and particulate) for all stations that include trace metals.
 * Macronutrients for all stations for all stations excluding station I008
 * Dissolved trace metals for all stations excluding station I008
 * Particulate trace metals for all stations excluding station I008

Permutation tests were run based on 999 permutations to evaluate the significance of the CCA axes and the significance of the environmental variables.


# CCA analysis
```{r setup, include=FALSE}

#rmarkdown options
knitr::opts_chunk$set(echo = TRUE,
                      results = "asis",
                      echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      error = FALSE)

options(knitr.duplicate.label = 'allow') #this will allow repeating chunknames. Needed as we repeat the same code for the analysis

# load libraries
library(readxl)
library(vegan)
library(dplyr)
library(tidyr)
library(corrplot)
library(ggplot2)
library(ggvegan)
library(pander)
library(knitr)
library(ggrepel)
library(kableExtra)
library(readr)

#load full dataset
WC17 <- read_csv(here::here("raw data/WC17_DataComp_update.csv")) #here:here is to get the project path
WC17stationlabels <- read_excel(here::here("raw data/WC17-CTD_Tables.xlsx"))
WC17stationlabels <- 
  WC17stationlabels %>%
  select(Station,Zone,`Station Label`)

WC17 <- WC17 %>%
        left_join(WC17stationlabels)%>%
        select(Cruise,Station,Zone,`Station Label`,everything())
        


```

\br
## Verifying assumptions: unimodality


```{r, fig.cap = "Density plots of CHEMTAX derived taxonomical groups' biomass (µg/L of Chl a)"}

WC17 %>%
        select(
          Diatoms:Prochlorococcus
        ) %>%
  tibble::rownames_to_column(var = "id") %>%
  pivot_longer(-id,
              names_to = "species",
              values_to = "concentration") %>%
  ggplot() +
  geom_density(aes(x=concentration))+
  facet_wrap(.~species,scales = "free")+
  xlab(bquote("CHEMTAX derived biomass (µg/L of Chl a)"))+
  ylab("Density")+
  theme_classic()


```
\br
From the density plots above, the phytoplankton biomass for the various taxonomic groups appear to be unimodal.

```{r, fig.cap = "Density distribution plots of environmental variables", fig.height = 12, fig.width = 12}

WC17_plot <- WC17 %>% 
        rename(
         "Nitrate (µmol/L)" = Nitrate,
         "Phosphate (µmol/L)" = `Phosphate` ,
         "Silica (µmol/L)"= `Silica`,
          "dFe (nmol kg-1)" = dFe,
          "dMn (nmol kg-1)" = dMn,
          "pMn (µmol kg-1)" = pMn,
          "dCo (pmol kg-1)" = dCo,
          "pCo (pmol kg-1)" = pCo,
          "dZn (nmol kg-1)" = dZn,
          "pZn (nmol kg-1)" = pZn,
          "dCd (nmol kg-1)" = dCd,
          "pCd (pmol kg-1)" = pCd,
          "dNi (nmol kg-1)" = dNi,
          "pNi (pmol kg-1)" = pNi,
          "dCu (nmol kg-1)" = dCu,
          "pCu (pmol kg-1)" = pCu,
          "P (nmol kg-1)" = P,
      "pAl (nmol kg-1)" = pAl,
      `Temp (°C)` = Temp,
      `Sal (psu)` = Sal
         )

WC17_plot %>%
          select(`Temp (°C)`,
                 `Sal (psu)`,
                 `Nitrate (µmol/L)`:`Silica (µmol/L)`,
                 `dFe (nmol kg-1)`:`pAl (nmol kg-1)`,
                  Phaeo_Chla) %>%
  tibble::rownames_to_column(var = "id") %>%
    pivot_longer(-id,
                names_to = "nutrient",
                values_to = "concentration") %>%
  ggplot() +
  geom_density(aes(x=concentration))+
  facet_wrap(.~nutrient,scales = "free")+
  theme_classic()+
  xlab("")+
  ylab("Density")


```

\br
The density plots above show that the nutrient distributions are mostly unimodal.


## CCA results
```{r}
### Rename variables
WC17 <- WC17 %>% 
        rename(
        `NO3` = "Nitrate",
         `PO4` = "Phosphate",
         `Si(OH)4` = "Silica"
                )

```
\br


```{r setupPlotObject}


plots <- NULL
```
### Analysis 1: All variables


```{r}

variables <- c("Temp", "Sal", 
               "NO3", "PO4", "Si(OH)4","dFe","dMn", "pMn", 
               "dCo", "pCo", "dZn", "pZn", "dCd", "pCd", 
               "dNi", "pNi", "dCu", "pCu")

stationsToInclude <- c("IO08", "IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")

variableDescription <- "Temperature, salinity, macronutrients, trace metals and phaeopigments:Tchla"

ccaFigName <- "allvariables"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out <- knit_expand(here::here('child.Rmd'), envir = env)

```
\br

`r paste(knit(text = out), collapse = '\n')`

### Analysis 2: Macronutrients


```{r}

variables <- c("NO3","PO4","Si(OH)4")
stationsToInclude <- c("IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "macronutrients"
ccaFigName <- "macronutrients"
corFigName <- paste("cor",ccaFigName,sep = "_")

env2 <- new.env()
out2 <- knit_expand(here::here('child.Rmd'), envir = env2)

```

`r paste(knit(text = out2), collapse = '\n')`

### Analysis 3: Macronutrients with STZ station


```{r}

variables <- c("NO3","PO4","Si(OH)4")
stationsToInclude <- c("IO08","IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "macronutrients with STZ"
ccaFigName <- "macronutrients_stz"
corFigName <- paste("cor",ccaFigName,sep = "_")

env2 <- new.env()
out2 <- knit_expand(here::here('child.Rmd'), envir = env2)

```


`r paste(knit(text = out2), collapse = '\n')`

### Analysis 4: macro and micro nutrients

```{r}

variables <- c(
               "NO3", "PO4", "Si(OH)4","dFe", "pFe","dMn", "pMn", 
               "dCo", "pCo", "dZn", "pZn", "dCd", "pCd", 
               "dNi", "pNi", "dCu", "pCu")
stationsToInclude <- c("IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "macronutrients and trace metals"
ccaFigName <- "macronutrients_and_tracemetals"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out3 <- knit_expand(here::here('child.Rmd'), envir = env)

```

`r paste(knit(text = out3), collapse = '\n')`

### Analysis 5: Dissolved trace metals 


```{r}

variables <- c("dFe", "dMn",  
               "dCo", "dZn", "dCd",  
               "dNi", "dCu")

stationsToInclude <- c("IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "dissolved trace metals"
ccaFigName <- "dissolved trace metals"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out4 <- knit_expand(here::here('child.Rmd'), envir = env)

```

`r paste(knit(text = out4), collapse = '\n')`
### Analysis 6: Particulate trace metals


```{r}

variables <- variables <- c("pMn","pCo", "pZn","pCd", "pNi", "pCu","pFe")
stationsToInclude <- c("IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "Particulate trace metals"
ccaFigName <- "particulate trace metals"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out5 <- knit_expand(here::here('child.Rmd'), envir = env)

```

`r paste(knit(text = out5), collapse = '\n')`

### Analysis 7: Dissolved trace metals including the STZ station


```{r}

variables <- c("dFe", "dMn",  
               "dCo", "dZn", "dCd",  
               "dNi", "dCu")

stationsToInclude <- c("IO08","IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "dissolved trace metals with STZ"
ccaFigName <- "dissolved trace metals_stz"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out6 <- knit_expand(here::here('child.Rmd'), envir = env)

```

`r paste(knit(text = out6), collapse = '\n')`


### Analysis 8: Particulate trace metals with STZ


```{r}

variables <- variables <- c("pMn","pCo", "pZn","pCd", "pNi", "pCu","pFe")
stationsToInclude <- c("IO08","IO07", "IO06", "IO05", "IO04", "IO03", "IO02", "IO01")
variableDescription <- "Particulate trace metals including the STZ station"
ccaFigName <- "particulate trace metals_stz"
corFigName <- paste("cor",ccaFigName,sep = "_")

env <- new.env()
out7 <- knit_expand(here::here('child.Rmd'), envir = env)

```


`r paste(knit(text = out7), collapse = '\n')`


```{r finalizing}

#saving the plots into a Rdata object that can be used in another script

save(plots,file = here::here("figures/ccaPlots.Rda"))

### Combining figures without having to run another script
#Example script to combine multiple plots using the package patchwork
library(ggplot2)
library(png)
library(patchwork) 
library(cowplot) ##can use cowplot as well to combine plots but here using to extract legend

#load plots from CCA analysis
load(here::here("figures/ccaPlots.Rda"))

p <- plots[["allvariables"]]

#can remove legend to add separately
legend <- ggpubr::get_legend(p, position = NULL)
legend<- ggpubr::as_ggplot(legend)
#plot annotation: https://patchwork.data-imaginist.com/articles/guides/annotation.html
plots_no_leg <- NULL
for(nm in names(plots)){
  plots_no_leg[[nm]] <- plots[[nm]] + theme(legend.position = "none",plot.margin = unit(c(0,0,0,0),"cm")) + coord_cartesian()
}

# Create the first four plots as a grid with labels
plot_grid <- plot_grid(
  plots_no_leg[["dissolved trace metals_stz"]], 
  plots_no_leg[["particulate trace metals_stz"]],
  plots_no_leg[["dissolved trace metals"]], 
  plots_no_leg[["particulate trace metals"]],
  labels = c("a)", "b)", "c)", "d)"),  # Custom labels
  label_size = 14,                     # Adjust label size if needed
  ncol = 2                             # Arrange in 2 columns
)+
  theme(plot.margin = unit(c(0.2,0.2,0.5,0.2),"cm"))

# Combine the grid with the legend below
combinedPlot <- plot_grid(
  plot_grid, 
  legend, 
  ncol = 1,                 # Stack grid and legend vertically
  rel_heights = c(0.9, 0.1) # Adjust relative heights of the components
) +
  theme(plot.background = element_rect(fill = "white", color = "white"),
panel.background = element_rect(fill = "white", color = "white")
)

# Display the combined plot
combinedPlot
ggsave(here::here("figures/figure5Combined.png"),
       height = 11,
       width = 8,
       dpi =300)


#### Macronutrients figure ###################
#load plots from CCA analysis
load(here::here("figures/ccaPlots.Rda"))

p <- plots[["macronutrients_stz"]]

#can remove legend to add separately
legend <- ggpubr::get_legend(p, position = NULL)
legend<- ggpubr::as_ggplot(legend)
#plot annotation: https://patchwork.data-imaginist.com/articles/guides/annotation.html
plots_no_leg <- NULL
for(nm in names(plots)){
  plots_no_leg[[nm]] <- plots[[nm]] + theme(legend.position = "none",plot.margin = unit(c(0,0,0,0),"cm")) + coord_cartesian()
}



# Create the first four plots as a grid with labels
plot_grid <- plot_grid(
  plots_no_leg[["macronutrients_stz"]],
  plots_no_leg[["macronutrients"]], 
  labels = c("a)", "b)"),  # Custom labels
  label_size = 14,                     # Adjust label size if needed
  ncol = 2                             # Arrange in 2 columns
)+
  theme(plot.margin = unit(c(0.2,0.2,0.5,0.2),"cm"))

# Combine the grid with the legend below
combinedPlot <- plot_grid(
  plot_grid, 
  legend, 
  ncol = 1,                 # Stack grid and legend vertically
  rel_heights = c(0.45, 0.1) # Adjust relative heights of the components
) +
  theme(plot.background = element_rect(fill = "white", color = "white"),
        panel.background = element_rect(fill = "white", color = "white")
  )

# Display the combined plot
combinedPlot
ggsave(here::here("figures/macronutrientsCombined.png"),
       height = 6,
       width = 8,
       dpi =300)

ggsave(here::here("figures/macronutrientsCombined.pdf"),
       height = 6,
       width = 8,
       dpi =300)
```
